Interactive effects of drought and habitat fragmentation on vital rates of an understory tropical plant

### 2020-11-18 analysis/paper/paper.Rmd
cleaned up and fleshed out outline of introduction

institute:
  - UF: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430
  - Columbia: Department of Ecology, Evolution and Environmental Biology, Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/ecology.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  forest fragmentation; vital rates; *Heliconia acuminata*; drought; ENSO; edge effects
highlights: |
  These are the highlights
---

@@

1.  Little is known about how organisms respond to multiple anthropogenic stresses simultaneously <!--# (see Lakeman-Fraser and Ewers 2014)  Might not be a good introduction if we can't do much with interactions. --> ... how multiple sources of biotic and abiotic stress at different scales affect demography. ... increasingly antrhropogenic...

    -   "multiple environmental factors" instead of "anthropogenic stresses"?

@@

2.  The Amazon Basin (or tropical lowland forests), an important place, has been fragmented, and although deforestation has slowed in recent years, it is continuing.

    1.  fragmentation is bad because loss of microclimate buffering, edge effects.

3.  Simultaneously, there is a drying trend in the Amazon and also increasing intensity and frequency of droughts caused by ENSO (wet-season) and sea surface warming (dry-season).

    1.  Drought is bad because...

4.  The effects of fragmentation and drought could be additive, or interactive. Could be both positive or negative depending on the vital rate.

    1.  The effects of drought could be more severe in fragments because of reduced microclimate buffering.

    2.  Alternatively, plants in fragmented habitats could be more resilient to drought because the individuals surviving in fragmented forests are already resilient in the face of a relatively dry and variable climate, even in non-drought years.

5.  In addition to variation among vital rates, responses to drought and fragmentation cuold depend on life history stage or plant size. Large established adults could repsond differently than seedlings. This is unknown because there are very few systems where we have data on entire life history in a climate change or fragmentation context.

    1.  Fragmentation affects plant size (in *Heliconia* [@brunaDemographicEffectsHabitat2005] and trees [@schwartz2019]). Smaller/larger plants might be more susceptible to drought. So differential effects of drought in fragments and continuous forest could be due to an interaction between drought and plant size.

6.  We investigated the effects of drought on the growth, survival, reproduction, and recruitment of an understory plant in an experimentally fragmented landscape.

    1.  Specifically, we want to know 1) Does drought have different impacts on plants in forest fragments compared to intact forest? 2) Do the magnitudes and directions of these effects differ among vital rates? 3) Do the effects of drought, fragmentation, and their interaction vary with plant size <!--# are there size dependent effects of drought and fragmentation? 4) are the effects of drought and fragmentation indpenedent, additive, or synergistic -->

# Notes for Introduction (first attempt at outline):

Habitat fragmentation is bad, pervasive, and getting worse

-   stats about global land use change <!--# start this broad or start with the Amazon? -->

-   stats about amazon deforestation [@malhi2008]

-   stats about how much land is near an edge (refs in Schwartz et al and betts 2019)

-   Why fragmentation bad

    -   microclimate changes [@ewersFragmentationImpairsMicroclimate2013]

    -   metapopulation and connectivity stuff (necessary?)

-   Plants in the tropics are less well adapted to fragmentation because lack of historical distrubance and less variable climate [@betts2019]

## Climate change and drought

-   The northern Amazon has been experiencing a drying trend since the mid 1970s and ensemble climate models predict substantial decreases in dry season precipitation in southern Amazonia for the 21st century [@malhi2008].

-   The El Niño--Southern Oscillation (ENSO) reduces wet-season rainfall in the Amazon <!--# is this correct? Or only in the south? -->. ENSO-induced droughts are predicted to increase in frequency and severity. [@caiIncreasingFrequencyExtreme2014].

-   Also sea surface temperature related drought. Warming of north Atlantic results longer and more intense dry season in southern and eastern Amazonia [@phillipsDroughtSensitivityAmazon2009; @malhi2008]

-   drought effects on plant vital rates

@@

-   [@zengCausesImpacts20052008]

## Interactions

-   However, our understanding of how climate change and habitat fragmentation might interact to impact organisms is limited.

-   [@schwartz2019] (mention how tree size classes affected differently, but we don't know if this is also true for understory herbaceous plants.

-   [@malhi2008]

-   The 1997--1998 ENSO drought caused greater tree mortality and increased leaf litter fall in experimentally fragmented forests compared to intact forest in the Amazon [@laurancePositiveFeedbacksForest2001].

-   Furthermore, the interactive effects of fragmentation and drought may be difficult to detect in long-lived species (e.g. Heliconia) [bruna paper that shows growth is reduced after 10 years slightly, but impacts reproduction]

## This experiment

-   The "fragmentation drought hypothesis" expects drought to have stronger impacts on forest fragments because of reduced microclimate buffering. Under this hypothesis we would expect drought to reduce growth, survival, reproduction and recruitment more in fragmented landscapes compared to intact forest.

-   It's also possible that plants in fragments have already been filtered such that the survivors are already adapted to more variable and desiccating conditions associated with forest edges [@schwartz2019]. This may result in drought having a less noticeable impact on plants in fragments, although we would expect fragmentation to still have an impact on survival, growth, and reproduction.

-   We set out to answer the following questions: 1) Does drought have different impacts on plants in forest fragments compared to intact forest? 2) Do the magnitudes and directions of these effects differ among vital rates? 3) Do the effects of drought, fragmentation, and their interaction vary with plant size [@tredennickSizebyenvironmentInteractionsNeglected2018] <!--# read this referrebce --> ?

# Background

@@

@@

>>>>

institute:
  - UF: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430
  - Columbia: Department of Ecology, Evolution and Environmental Biology, Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
linkcolor: blue
bibliography: references.bib
csl: "../templates/ecology.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  forest fragmentation; vital rates; *Heliconia acuminata*; drought; ENSO; edge effects
highlights: |
  These are the highlights
---

@@

1.  Organisms experience multiple sources of natural and anthropogenic stress simultaneously. However, the impacts of multiple sources of stress on population dynamics are not well understood [@tyeDemographicMenageTrois2016].

@@

2.  As a result of intense deforestation, tropical lowland forests are becoming an increasingly fragmented landscape. This is especially true in the Amazon basin where \_\_\_\_ km^2^ forest are within 100m of a forest edge <!--# citation -->.

    -   also cite: Kapos 1989, Lovejoy et al. 1986, Kapos et al. 1997?

    -   This can result in ...

3.  Simultaneously, these fragmented habitats are experiencing increased drought stress as a result of climate change. The northern Amazon has been experiencing a drying trend since the mid 1970s and ensemble climate models predict substantial decreases in dry season precipitation in southern Amazonia for the 21st century [@malhi2008]. Additionally, El Niño Southern Oscillation (ENSO)-induced droughts are predicted to increase in frequency and severity [@caiIncreasingFrequencyExtreme2014].

@@

    -   Drought is bad because...

    -   Combined environmental stresses, such as drought and fragmentation, have the potential to exacerbate or dampen effects on population dynamics [@tyeDemographicMenageTrois2016].

4.  Plant populations may be more susceptible to drought in fragmented forests. Fragmentation results in changes in forest microclimate that could exacerbate the effects of drought on plant populations. Near edges, tropical forests experience increased solar radiation, increased temperature, and decreased humidity and soil moisture due to reduced microclimate buffering [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999].

    -   Trees near forest edges experience increased mortality after droughts compared to forest interiors [@schwartz2019; @lauranceEffectsStrongDrought2001].

    -    Increased mortality after droughts disproportionately affects large trees.

5.  Alternatively, plant individuals that are more susceptible to desiccation may not have survived fragmentation, resulting in populations in fragments which are more resilient to drought.

    -   [@betts2019].

6.  <!--# I'm now thinking this doesn't need to be a separate paragraph, but included in #4.  Need to do some more reading about this though. -->In addition to variation among vital rates, responses to drought and fragmentation could depend on life history stage or plant size. Large established adults could respond differently than seedlings. This is unknown because there are very few systems where we have data on entire life history in a climate change or fragmentation context.

    -   Fragmentation affects plant size (in *Heliconia* [@brunaDemographicEffectsHabitat2005] and trees [@schwartz2019]). Smaller/larger plants might be more susceptible to drought. So differential effects of drought in fragments and continuous forest could be due to an interaction between drought and plant size.

7.  We investigated the effects of drought on the growth, survival, reproduction, and recruitment of an understory plant in an experimentally fragmented landscape. Specifically, we want to know 1) Does drought increase or decrease the growth, survival, and fertility rates of plant populations in continuous forest? 2) Are the effects of drought on the vital rates of populations in fragments similar in direction and magnitude to those in continuous forest? 3) Are the effects of drought and fragmentation on vital rates consistent across life-history stages?

@@

<!--# Emilio, feel free to edit -->

@@

<!--# Emilio: please edit -->
